Cultural Evolution of Conservation Behavior at the Intersection of Working

### 2020-10-08 DissertationProposal/DissertationProposal.Rmd
small change
text edits for clarity

\newpage
# Summary
Area-based conservation is critical for stemming global biodiversity loss. As such, each year billions of dollars are spent to protect key areas. It is common, however, for these areas to fail to reach their stated conservation goals. The most chronic causes of conservation project failure are the human dimensions. Conversely, historic conservation successes are also attributed to widespread human adoption of conservation behaviors. Research examining how and why conservation practices are adopted tends to suffer from a bias toward the reporting of successful projects and highly context dependent drivers and outcomes. This dissertation seeks to demonstrate and better understand the underlying mechanisms which underpin social adoption of conservation practices in and around protected areas. Toward this aim, this project will combine insights from cultural group selection theory with empirical research on Pemba Island, Tanzania to achieve three primary objectives.

1. Develop a dynamic model to generate predictions regarding if and how conservation norms can emerge and persist via competition between groups.

2. Empirically test theoretical predictions of group conservation adoption as a function of group competition and individual costs and benefits.

3. Using the ongoing spread of an improved efficiency cookstove design, specifically examine how the nature of interpersonal relationships may modify conservation behaviors as they spread within groups.

@@

## Primer on cultural group selection
The largely untested theory of cultural group selection poses a mechanism for the emergence and spread of individually costly yet group beneficial behaviors such as warfare or religious donations [@boyd_transmission_2010]. Cultural group selection suggests that groups who develop individually costly yet group beneficial norms will outcompete groups who do not. As an individual’s behavior is flexible, individuals will either migrate from less to more successful groups or directly copy the cooperative behavior of others in visibly successful cooperative groups [@henrich_cultural_2004]. This change in the frequency of behavioral traits is analogous to the perpetuation of beneficial alleles in genetic natural selection. In these contexts, selective pressures still push individuals towards selfish behaviors like freeriding, but cooperative norms will emerge if group selection pressures outweigh those of individual selection overall [@mcelreath_mathematical_2008; @okasha_multilevel_2004; @price1972extension].

@@

This project aims to logically test the applicability of cultural group selection to explain the emergence and spread of protected area adoption. In this chapter I will develop a dynamic, agent-based simulation of individuals harvesting resources from a landscape under a variety of environmental constraints. I will then use this simulation to test under what circumstances adherence to protected area regulations emerges as a group norm within solitary and multiple, competing groups. This model will generate qualitative predictions about system outcomes given a very simple set of assumptions.

@@

### Model Description
In the most simple scenario, the agent based model represents a simplistic world where there is one homogeneous group of agents who harvest resources from a landscape. This landscape is divided into two types, working and protected. Resources move between these patches at each time step with some probability. During initialization, the protected and working landscapes are both at an equal, designated percent carrying capacity. At each time step, agents aim to maximize their harvest up until a designated harvest max. During initialization, agents are assigned as either obligatory cooperators or as having the ability to defect to a small degree, spending a small portion of their time harvesting from the protected landscape.

@@

The environment also responds to agent behavior. At every time step, both the protected and working landscapes lose exactly the number of resources that agents harvest from them. The areas then regenerate resources according to the regeneration rate per time step and lastly, the resources move from one area to the other with the probability specified during model initiation.

@@

@@

```{r conceptDiagram, echo=FALSE, fig.cap="Concept diagram of agent-based model scheduling under single group dynamics", out.width = "50%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch1/ConceptDiagram.png")
```

@@

@@

The key research questions for this project will be assessed using multilevel regression operationalized in a Bayesian context. I will compare standardized parameter estimates, to determine which factors (geographic distance or social relatedness) most strongly contribute to copying fidelity in this study system. Similar analyses will be used to determine which factors affect cookstove primary outcome variables (fuel efficiency and particulate matter production).

>>>>

\newpage
# Summary
Area-based conservation is critical for stemming global biodiversity loss. As such, each year billions of dollars are spent to protect key areas. It is common, however, for protected areas to fail to reach their stated conservation goals. The most chronic cause of conservation failure is a lack of local support and adoption of mandates. Conversely, historic protected area successes are also attributed to widespread adoption of conservation behaviors. Research examining how and why protected areas are adopted tends to suffer from a bias toward the reporting of successful projects and highly context dependent drivers and outcomes. This dissertation seeks to demonstrate and better understand the underlying mechanisms which underpin local adoption of conservation practices in and around protected areas. Toward this aim, this project will combine insights from cultural group selection theory with empirical research on Pemba Island, Tanzania to achieve three primary objectives.

1. Develop a dynamic, theoretical model to generate predictions regarding if and how group-level adoption of protected areas can emerge and persist via competition between groups.

2. Empirically test theoretical predictions of protected area adoption as a function of group competition and perceived costs and benefits to local community members.

3. Using the ongoing spread of an improved efficiency cookstove design, examine how the nature of interpersonal relationships may modify conservation behaviors as they spread within groups.

@@

## Primer on cultural group selection
The largely untested theory of cultural group selection poses a mechanism for the emergence and spread of individually costly yet group beneficial behaviors such as warfare or religious donations [@boyd_transmission_2010]. Cultural group selection suggests that groups who develop individually costly yet group beneficial norms will outcompete groups who do not. As an individual’s behavior is flexible, individuals may migrate from less to more successful groups or directly copy the behaviors they observe in visibly successful cooperative groups [@henrich_cultural_2004]. This change in the frequency of behavioral traits is analogous to the perpetuation of beneficial alleles in genetic natural selection. In these contexts, selective pressures still push individuals towards selfish behaviors like freeriding, but cooperative norms will emerge if group selection pressures outweigh those of individual selection overall [@mcelreath_mathematical_2008; @okasha_multilevel_2004; @price1972extension].

@@

This project aims to logically test the applicability of cultural group selection to conservation. Specifically, this research will address how behavior-environment feedbacks and resource mobility and regeneration rates affect the emergence and spread of protected area adoption. In this chapter I will develop a dynamic, agent-based simulation of individuals harvesting resources from a landscape under a variety of environmental constraints. I will then use this simulation to test under what circumstances adherence to protected area regulations emerges as a group norm within solitary and multiple, competing groups. This model will generate qualitative predictions about system outcomes given a very simple set of assumptions.

@@

### Model Description

*Single group setup and basic function*

In the most simple scenario, the agent-based model represents a simplistic world where there is one homogeneous group of agents who harvest resources from a landscape. This landscape is divided into two types, working and protected. Resources move between these patches at each time step with some probability. During initialization, the protected and working landscapes are both at an equal, designated percent carrying capacity. At each time step, agents aim to maximize their harvest up until a designated harvest max. During initialization, agents are assigned as either obligatory cooperators or as having the ability to defect to a small degree, spending a small portion of their time harvesting from the protected landscape.

@@

The environment also responds to agent behavior. At every time step, both the protected and working landscapes lose exactly the number of resources that agents harvest from them. The areas then regenerate resources according to the regeneration rate per time step and lastly, the resources move from one area to the other with the probability specified during model initialization.

@@

*Key processes*

@@

```{r conceptDiagram, echo=FALSE, fig.cap="Concept diagram of agent-based model scheduling under single group dynamics", out.width = "50%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch1/ConceptDiagram.png")
```
*Multiple competing groups*

@@

*Outcomes*

@@

The key research questions for this project will be assessed using multilevel regression operationalized in a Bayesian context. I will compare standardized parameter estimates, to determine which factors (geographic distance or social relatedness) most strongly contribute to copying fidelity in this study system. Similar analyses will be used to determine which factors, if any, are strongly associated with stove fuel efficiency.
